Best Practices for Single Alchemical Free Energy Calculations [Article v\versionnumber]

### 2025-04-29 paper/manuscript.tex
Quick and dirty first complete version; now wait for feedback  On branch sboresch
 Changes to be committed:
	modified:   manuscript.tex

The influence from an incorrect treatment of dummy atoms (e.g., leaving all bonded terms intact) is difficult to estimate, though it may be small in many cases \cite{Fleck2021}. Nevertheless, some guidelines should be followed:
\begin{itemize}
  \item Obviously, a system with and without dummy atoms should have the same energetic minimum. Note that in larger groups of dummy atoms, redundant bonded terms involving only dummy atoms will be essential to avoid flapping, and the energy of the optimized geometry may be different from that of the physical system without dummy atoms. Thus, a \underline{necessary} condition is as follows: Optimize the geometry of the physical system (no dummy atoms); the final energy is the reference value. Optimize the geometry of the same molecule in the presence of dummy atoms; the final energy may well be different from the reference energy. Now delete the dummy atoms and recompute the energy without further minimization. The energy after the deletion of the dummy atoms must be identical to the reference energy. 
\item The above is only a \underline{necessary} criterion, and the dummy atoms may still influence the dynamical behavior of the physical molecule. Furthermore, overzealous removal of redundant energy terms may result in flapping of the dummy atoms. If one suspects this to be the case, then a short simulation in the gas phase should reveal flapping. Unfortunately, while straightforward in principle, both checks described are extremely hard to automate, which is essential in large RFEs campaigns. 
\item The correct treatment of dummy atoms can be validated by closing a suitable thermodynamic cycle.  E.g., Fleck et al. calculated ASFEs for each of their endstates, as well as the RSFE between them. In all cases when dummy atoms were treated according to best practices, the cycle closure error was well within the statistical uncertainty, which was well below 0.1 kcal/mol. A related, earlier study by Loeffler et al. also reported ASFEs and RSFEs. If one attempts to close the cycle based on the numbers reported there, cycle closure error was nonnegligible in some cases. [Loeffler2017]
\item While solvation free energies can nowadays be computed quite fast, their systematic calculation would nevertheless be too costly. The role of solvent can be replaced by some external restraint, distorting the average geometry of the physical molecule. If one computes the free energy cost of applying the restraint with and without dummy atoms, the two numbers must agree within statistical significance. While much faster than full solvation free energy calculations, applying a restraint in the gas phase requires careful thermalization of the simulation system, and automation is difficult.
  \end{itemize}
Further work automating the correct setup and treatment of dummy atoms is necessary.

>>>>

The influence from an incorrect treatment of dummy atoms (e.g., leaving all bonded terms intact) is difficult to estimate, though it may be small in many cases \cite{Fleck_2021}. Nevertheless, some guidelines should be followed:
\begin{itemize}
  \item Obviously, a system with and without dummy atoms should have the same energetic minimum. Note that in larger groups of dummy atoms, redundant bonded terms involving only dummy atoms will be essential to avoid flapping, and the energy of the optimized geometry may be different from that of the physical system without dummy atoms. Thus, a \underline{necessary} condition is as follows: Optimize the geometry of the physical system (no dummy atoms); the final energy is the reference value. Optimize the geometry of the same molecule in the presence of dummy atoms; the final energy may well be different from the reference energy. Now delete the dummy atoms and recompute the energy \underline{without} further minimization. The energy after the deletion of the dummy atoms must be identical to the reference energy. 
\item The above is only a \underline{necessary} criterion, and the dummy atoms may still influence the dynamical behavior of the physical molecule. Furthermore, overzealous removal of redundant energy terms may result in flapping of the dummy atoms. If one suspects the latter, then a short simulation in the gas phase should reveal flapping. Unfortunately, while straightforward in principle, both checks described are hard to automate, which is essential in large RFEs campaigns. 
\item The correct treatment of dummy atoms can be validated by closing a suitable thermodynamic cycle.  E.g., Fleck et al.\ calculated ASFEs for each of their endstates, as well as the RSFE between them \cite{Fleck_2021}. In all cases when dummy atoms were treated according to best practices, the cycle closure error was well within the statistical uncertainty, which was well below 0.1 kcal/mol. A related, earlier study by Loeffler et al. also reported ASFEs and RSFEs. If one attempts to close the cycle based on the numbers reported there, cycle closure error was nonnegligible in some cases \cite{loeffler2018reproducibility}. 
\item While solvation free energies can nowadays be computed quite fast, their systematic calculation would nevertheless be too costly. The role of solvent can be replaced by some external restraint, distorting the average geometry of the physical molecule. If one computes the free energy cost of applying the restraint with and without dummy atoms, the two numbers must agree within statistical significance. While much faster than full solvation free energy calculations, applying a restraint in the gas phase requires careful thermalization of the simulation system, and automation is difficult.
  \end{itemize}
%Further work automating the correct setup and treatment of dummy atoms is necessary.
